Figures

### 2025-02-04 Figures/PaperFiguresextended.Rmd
Added code for DEG and CellChat analysis

Figure 1c
UMAP of the entire endometrium single-cell atlas from women with and without endometriosis, colored by cell types (n=63).
```{r}
# Define order of cell types
order <- EndoAtlas_meta %>%
  select(AnnotationMain, AnnotationRefined) %>%
  filter(!is.na(AnnotationMain)) %>%
  distinct() %>%
  arrange(AnnotationRefined) %>%
  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
  arrange(AnnotationMain)
EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
# DimPlot
Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
Fig1c
# Save DimPlot
ggsave(filename = paste0(dir_out, 'Fig1c.pdf'),
       plot = Fig1c,
       width =20, height=9)

@@

Figure 1d
Main cell type frequencies per patient (n=60, box plot, left) and entire dataset (bar plot, middle), with mesenchymal cells being the most abundant cell type. The overall cell number distribution between non-ENDO, mild-ENDO and severe-ENDO is balanced (bar plot, right).
```{r plotting of cell types per sample}
## Fig1d_1 Cells per patient
# Boxplot
Fig1d_1 <- EndoAtlas_meta %>%
  group_by(AnnotationMain) %>%
  summarise(row_count = n()) %>% 
  ggplot(aes(x = AnnotationMain, y = row_count)) +
  geom_col() +
  coord_flip()
Fig1d_1
# Save Boxplot
ggsave(paste0(dir_out, "Fig1d_1.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

## Fig1d_2 Cells per main cell type
# Barplot
Fig1d_2 <- EndoAtlas_meta %>%
  group_by(sample, AnnotationMain) %>%
  summarise(row_count = n()) %>%
  ggplot(aes(x = AnnotationMain, y = row_count)) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
Fig1d_2
# Save Barplot
ggsave(paste0(dir_out, "Fig1d_2.pdf"),
         plot = last_plot(),
         width = 15, height = 6)

##Fig1d_3 cell number distribution between non-ENDO, mild-ENDO and severe-ENDO
# Barplot
Fig1d_3 <- EndoAtlas_meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ggplot(aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
Fig1d_3
# Save Barplot
ggsave(paste0(dir_out, "Fig1d_3.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

@@

Figure 1e
Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.

CODE running, need to extend the PCA analysis folder
dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"
```{r}
## Figure 1e
# Load data
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
# Plot and save eigencorplot
pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
)
dev.off()
```

Figure 2a
Scores plot of the first two principal components from PCA of sample-wise whole transcriptome.
```{r}
## Figure 2a
# Load data
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
# biplot
biplot(p,
    colby = 'MenstrualCyclePhase',
    # ellipse config
      ellipse = TRUE,
      ellipseType = 't',
      ellipseLevel = 0.90,
      ellipseFill = TRUE,
      ellipseAlpha = 1/4,
      ellipseLineSize = 1.0,
    labSize = 0,drawConnectors = FALSE,
    legendPosition = 'top', legendLabSize = 16, legendIconSize = 8.0)
```

Figure 2b
UMAPs of the entire dataset (left) and the re-integrated epithelial cells (right), colored by menstrual cycle phase, show a clear visual separation between proliferative/periovulatory, early-, mid-, and late-secretory samples, particularly in mesenchymal (endometrial fibroblasts, C7 fibroblasts, and mural cells), epithelial (glandular), and endothelial cells.
```{r}
## Figure 2b entire atlas
#DimPlot
Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
Fig2b_1
#Save DimPlot
ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
       plot = Fig2b_1,
       width =20, height=9)

## Figure 2b epithelial cells
#DimPlot
Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
Fig2b_2
#Save DimPlot
ggsave(filename = paste0(dir_out, 'Fig2b_2.pdf'),
       plot = Fig2b_2,
       width =20, height=9)

@@

Figure 2c
UMAP of the re-integrated epithelial cells, colored by pseudo time prediction from monocle3.
```{r}
# Figure 2c epithelial cells
epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))

# DimPlot of epithelial cells with pseudotime
Fig2c <- plot_cells(epithelial_cells_monocle3, color_cells_by = "pseudotime",
                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
                     graph_label_size=3, cell_size = 1)
Fig2c
# Save DimPlot
ggsave(filename = paste0(dir_out, 'Fig2c.pdf'),
       plot = Fig2c,
       width =20, height=9)
```

Figure 2d
Heatmap of selected marker gene expression with high log fold changes over the key menstrual cycle phases (proliferative, periovulatory, early-, mid-, and late-secretory).
```{r}
#Remove samples 65 and 105, since these samples are just in between menstrual cycle phases and cannot be clearly assigned
Idents(EndoAtlas) <- EndoAtlas$sample
EndoAtlas_filtereds65s105 <- subset(EndoAtlas, subset = sample %in% setdiff(unique(EndoAtlas$sample), c("sample65", "sample105")))
Idents(EndoAtlas_filtereds65s105) <- EndoAtlas_filtereds65s105$AnnotationMain

#### Plot heatmaps for main cell type specific markers for the different menstrual cycle phases. Compare also to Wang et al 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)

@@

Figure 2e
Boxplots of sample-wise mesenchymal and epithelial cell frequency as a percentage of total cells, split by menstrual cycle phase (stringent patient exclusion criteria, 32 ENDO, and 22 non-ENDO samples).
```{r plotting of cell types per sample}
##Boxplots of main cell types cell frequency as a percentage of total cells, split by menstrual cycle phase and endometriosis status

#Prepare data for boxplot
df_Fig2e <- EndoAtlas_meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main, EndometriosisStatus) %>%
  summarise(row_count = n()) %>%
  ungroup() %>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample) %>%
  mutate(AnnotationMain_perc = row_count / sum(row_count) * 100) %>%
  ungroup()

#BoxPlots percentage, EndometriosisStatus 
Fig2e <- ggplot(df_Fig2e, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge()) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~AnnotationMain,scales = "free", ncol = 6)
ggsave(paste0(dir_out, "Fig2e_boxplot.pdf"),
  plot = Fig2e,
  width = 12, height = 4.4)

###Significance testing CTL vs Endo
# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
df_Fig2e_ttest_EndometriosisStatus <- df_Fig2e %>%
  group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#epithelial, proliferative just not significant (p.value = 0.0898)

###Significance testing prolif vs periov vs sec, CTL + Endo
## Perform t-test Prolif vs Peri for each cell type
df_Fig2e_ttest_ProlifPeri <- df_Fig2e %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#none significant
## Perform t-test Prolif vs Sec for each cell type
df_Fig2e_ttest_ProlifSec <- df_Fig2e %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00977) and str (0.0255) significant

## Perform t-test periovulatory vs sec for each cell type
df_Fig2e_ttest_PeriSec <- df_Fig2e %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00794) and str (0.0137) significant
```

Make better comments! xxx
Figure 2f
Boxplot of sample-wise secretory mural cell frequency as a percentage of total cells.
```{r plotting of cell types per sample}
#Early and Mid Secretory Prv_VSMCs 
df_Fig2f <- EndoAtlas_meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
df_Fig2f <- left_join(df_Fig2f, dplyr::distinct(EndoAtlas_meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- df_Fig2f %>%
  filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
  dplyr::filter(MenstrualCyclePhase %in% c("proliferative","periovulatory","secretory_mid","secretory_early")) %>%
  group_by(sample, EndometriosisStatus, MenstrualCyclePhase_main) %>%
  summarize(
    total_counts_MenstrualCyclePhase_main = sum(AnnotationRefined_counts, na.rm = TRUE),
    total_perc_MenstrualCyclePhase_main = sum(AnnotationRefined_perc, na.rm = TRUE)
  ) %>%
  ungroup()

@@

#Multiple testing correction
p.adjust(c(0.0261), method = "BH", n = length(3)) #Value does not change
p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change
```

Figure 3a and 3b
3a Barplot showing number of unique differentially expressed genes (DEGs, adjusted p-value < 0.05, average expression > 18.53202) between non-ENDO (n = 11) and all-ENDO (n = 12), mild-ENDO (n = 4) and severe-ENDO (n = 8) samples within immune, endothelial, epithelial, and mesenchymal cell types, analyzed by Muscat.
3b Selected terms enriched in the cluster- and stage-wise functional enrichment analysis of DEGs, with term sizes between 25 and 750.
```{r}

@@

xxx
Figure 3c-h
(c) Selected signaling pathways with overall enriched activity in ENDO (n = 12) in comparison to non-ENDO (n = 11).
(d) Circle plot of TNF signaling pathway between all cell types in ENDO (left, n = 12) and non-ENDO (right, n = 11).
(e-h) Bubble plots of selected communications with significant interaction probabilities (unadjusted p-value < 0.01) within the ICAM (e), CXCL (f), IGFBP (g) and VEGF (h) signaling pathways for ENDO (n = 12) and non-ENDO (n = 11). Further p-value adjustment is not recommended by the CellChat author. 
```{r}

@@

Supplementary Data Figure 1a and 1b
Violin plots illustrating counts and features from RNA assay (a) and SCT assay (b), show no discernable differences between ENDO and non-ENDO.
```{r VlnPlots of nCounts and nFeatures}
#VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"
VlnPlot(EndoAtlas,
  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"),
  group.by='EndometriosisStatus', ncol = 1, pt.size=0)
# Save VlnPlots
ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
  plot = last_plot(),
  width = 4, height = 13)
```

Supplementary Data Figure 1c and d
UMAPs of the entire endometrium single-cell atlas, colored by menstrual cycle day (c) and progesterone in serum (d), show a clear visual separation of cells by menstrual cycle day and serum progesterone concentration. 
```{r}
## SFig1c
# Featureplot
SFig1c <- FeaturePlot(EndoAtlas, features = "CycleDay", raster = FALSE)
SFig1c
# Save Featureplot
ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
       plot = SFig1c,
       width =20, height=9)

## SFig1d
# Featureplot
SFig1d <- FeaturePlot(EndoAtlas, features = "ProgesteroneSerum", raster = FALSE)
SFig1d
# Save Featureplot
ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
       plot = SFig1d,
       width =20, height=9)

@@

xxx double-check code
Supplementary Data Figure 1e
Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the proliferative samples with strict exclusion criteria with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.
```{r}
dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")

@@

Supplementary Data Figure 1f
Eigencor plots showing the effect of different co-variants of the atlas dataset on principal components (PCs) of the main 5 cell types (epithelial, mesenchymal, endothelial, myeloid, lymphoid) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure. 
"/home/common/data/output/projects/ENDO/E044/A082/SCTassay/"
```{r}

@@

Supplementary Data Figure 2
Expression of the marker genes for the different cell types corresponds to the cell type-specific expression shown in Tan et al. 2022 (DOI: 10.1038/s41556-022-00961-5).
```{refined cluster markers}

@@

Supplementary Figure 3a
Scores plot of the first two principal components PC1 and PC2 from sample-wise whole transcriptome PCA of the SCT assay, colored by endometriosis status.
```{r}
# Load data
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
# biplot
SFig3a <- biplot(p,
    colby = 'EndometriosisStatus',
    labSize = 0,drawConnectors = FALSE, 
    hline = 0, vline = 0,
    legendPosition = 'right')
ggsave(plot = SFig3a, "./SFig3a.pdf", width = 7, height = 7)
```

Supplementary Figure 3b and e
UMAP of re-integrated epithelial cells colored by menstrual cycle day and menstrual cycle phase Wang
```{r}
epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))

@@

xxxsee above for object name in GEO
Supplementary Data Figure 3c
Look in IBU folder for object with trajectory line or A063/A066
```{r}
#Figure S3c epithelial cells
FeaturePlot
Fig3c <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCycleDay", raster = FALSE)
Fig3c
ggsave(filename = paste0(dir_out, 'Fig3c.pdf'),
       plot = Fig3c,
       width =20, height=9)
```

>>>>

Figure 1c
CODE OK
```{r}

order <- EndoAtlas_meta %>%
  select(AnnotationMain, AnnotationRefined) %>%
  filter(!is.na(AnnotationMain)) %>%
  distinct() %>%
  arrange(AnnotationRefined) %>%
  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
  arrange(AnnotationMain)
EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
Fig1c
ggsave(filename = paste0(dir_out, 'Fig1c.pdf'),
       plot = Fig1c,
       width =20, height=9)

@@

Figure 1d
CODE OK
```{r plotting of cell types per sample}
#Fig1d_1
Fig1d_1 <- EndoAtlas_meta %>%
  group_by(AnnotationMain) %>%
  summarise(row_count = n()) %>% 
  ggplot(aes(x = AnnotationMain, y = row_count)) +
  geom_col() +
  coord_flip()
Fig1d_1
ggsave(paste0(dir_out, "Fig1d_1.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

#Fig1d_2
Fig1d_2 <- EndoAtlas_meta %>%
  group_by(sample, AnnotationMain) %>%
  summarise(row_count = n()) %>%
  ggplot(aes(x = AnnotationMain, y = row_count)) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
Fig1d_2
ggsave(paste0(dir_out, "Fig1d_2.pdf"),
         plot = last_plot(),
         width = 15, height = 6)

#Fig1d_3 
Fig1d_3 <- EndoAtlas_meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ggplot(aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
Fig1d_3
ggsave(paste0(dir_out, "Fig1d_3.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

@@

Figure 1e, 2a, S3a
CODE running, need to extend the PCA analysis folder
dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"

```{r}
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
#Figure 1e
pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
)
dev.off()

#Figure 2a
biplot(p,
    colby = 'MenstrualCyclePhase',
    # ellipse config
      ellipse = TRUE,
      ellipseType = 't',
      ellipseLevel = 0.90,
      ellipseFill = TRUE,
      ellipseAlpha = 1/4,
      ellipseLineSize = 1.0,
    labSize = 0,drawConnectors = FALSE,
    legendPosition = 'top', legendLabSize = 16, legendIconSize = 8.0)

#Supplementary Data Figure 3a
biplot(p,
    colby = 'EndometriosisStatus',
    labSize = 0,drawConnectors = FALSE, 
    hline = 0, vline = 0,
    legendPosition = 'right')
```

Figure 2b
CODE OK
```{r}
#Figure 2b entire atlas
Fig2b_1 <- DimPlot(EndoAtlas, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
Fig2b_1
ggsave(filename = paste0(dir_out, 'Fig2b_1.pdf'),
       plot = Fig2b_1,
       width =20, height=9)

#Figure 2b epithelial cells
Fig2b_2 <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCyclePhase", raster = FALSE)
Fig2b_2
ggsave(filename = paste0(dir_out, 'Fig2b_2.pdf'),
       plot = Fig2b_2,
       width =20, height=9)

@@

Figure 2c
Look in IBU folder for object with trajectory line or A063/A066
```{r}
#Figure 2c epithelial cells
epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))

#Pseudotime
Fig2c <- plot_cells(epithelial_cells_monocle3, color_cells_by = "pseudotime",
                     label_cell_groups=FALSE, label_leaves=FALSE, label_branch_points=FALSE,
                     graph_label_size=3, cell_size = 1)
Fig2c
ggsave(filename = paste0(dir_out, 'Fig2c.pdf'),
       plot = Fig2c,
       width =20, height=9)
```

Figure 2d
CODE OK
```{r}
#Remove samples 65 and 105, since these samples are just in between menstrual cycle phases and cannot be clearly assigned
Idents(EndoAtlas) <- EndoAtlas$sample
EndoAtlas_filtereds65s105 <- subset(EndoAtlas, subset = sample %in% setdiff(unique(EndoAtlas$sample), c("sample65", "sample105")))
Idents(EndoAtlas_filtereds65s105) <- EndoAtlas_filtereds65s105$AnnotationMain

#### Plot heatmaps for main cell type specific markers for the different menstrual cycle phases. Compare also Wang et al 2020 marker genes (Wang et al. 2020, DOI: 10.1038/s41591-020-1040-z)

@@

Figure 2e
CODE OK
```{r plotting of cell types per sample}
##Boxplots of main cell types cell frequency as a percentage of total cells, split by menstrual cycle phase and endometriosis status

#Prepare data for boxplot
df2 <- EndoAtlas_meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main, EndometriosisStatus) %>%
  summarise(row_count = n()) %>%
  ungroup() %>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample) %>%
  mutate(AnnotationMain_perc = row_count / sum(row_count) * 100) %>%
  ungroup()

#BoxPlots percentage, EndometriosisStatus 
ggplot(df2, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge()) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~AnnotationMain,scales = "free", ncol = 6)
ggsave(paste0(dir_out, "Fig2e_boxplot.pdf"),
  plot = last_plot(),
  width = 12, height = 4.4)

###Significance testing CTL vs Endo
# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
df2_ttest_EndometriosisStatus <- df2 %>%
  group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#epithelial, proliferative just not significant (p.value = 0.0898)

###Significance testing prolif vs periov vs sec, CTL + Endo
## Perform t-test Prolif vs Peri for each cell type
df2_ttest_ProlifPeri <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#none significant
## Perform t-test Prolif vs Sec for each cell type
df2_ttest_ProlifSec <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00977) and str (0.0255) significant

## Perform t-test periovulatory vs sec for each cell type
df2_ttest_PeriSec <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00794) and str (0.0137) significant
```

Figure 2f
CODE OK
```{r plotting of cell types per sample}
#Early and Mid Secretory Prv_VSMCs 
Fig2f <- EndoAtlas_meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
Fig2f <- left_join(Fig2f, dplyr::distinct(EndoAtlas_meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- Fig2f %>%
  filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
  dplyr::filter(MenstrualCyclePhase %in% c("proliferative","periovulatory","secretory_mid","secretory_early")) %>%
  group_by(sample, EndometriosisStatus, MenstrualCyclePhase_main) %>%
  summarize(
    total_counts_MenstrualCyclePhase_main = sum(AnnotationRefined_counts, na.rm = TRUE),
    total_perc_MenstrualCyclePhase_main = sum(AnnotationRefined_perc, na.rm = TRUE)
  ) %>%
  ungroup()

@@

#Multiple testing correction
p.adjust(c(0.0261), method = "BH", n = length(3)) #Value does not change
p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change

Figure 3a and 3b
```{r}

@@

Figure 3c-h
```{r}

@@

Supplementary Data Figure 1a and 1b
CODE OK
```{r VlnPlots of nCounts and nFeatures}
##VlnPlots for "nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"
VlnPlot(EndoAtlas,
  features = c("nFeature_RNA", "nCount_RNA","nCount_SCT","nFeature_SCT"),
  group.by='EndometriosisStatus', ncol = 1, pt.size=0)
ggsave(paste0(dir_out, "VlnPlot_GroupByEndometriosisStatus_nFeature_nCounts_RNA_SCT_percMT.pdf"),
  plot = last_plot(),
  width = 4, height = 13)

@@

```{r different counts and means/medians}
#Percentage of main 5
table <- as.data.frame(table(EndoAtlas_meta$AnnotationMain))
table$Percentage <- round((table$Freq / sum(table$Freq)) * 100, 1)

#Number of cells for CTL and Endo
table(EndoAtlas_meta$EndometriosisStatus)

#Number of patients for CTL and Endo
EndoAtlas_meta %>% select(sample,EndometriosisStatus) %>%
  distinct() %>%
  group_by(EndometriosisStatus) %>%
  summarize(number_of_rows = n())

#mean number of cells per sample
mean(table(EndoAtlas_meta$sample))

#median count/feature
median_summary <- EndoAtlas_meta %>%
  select(nCount_RNA, nCount_SCT, nFeature_RNA, nFeature_SCT) %>%
  summarize(
    median_nCount_RNA = median(nCount_RNA),
    median_nCount_SCT = median(nCount_SCT),
    median_nFeature_RNA = median(nFeature_RNA),
    median_nFeature_SCT = median(nFeature_SCT)
  )
```

Supplementary Data Figure 1c
CODE OK
```{r}
SFig1c <- FeaturePlot(EndoAtlas, features = "CycleDay", raster = FALSE)
SFig1c
ggsave(filename = paste0(dir_out, 'SFig1c.pdf'),
       plot = SFig1c,
       width =20, height=9)
```

Supplementary Data Figure 1d
CODE OK
```{r}
SFig1d <- FeaturePlot(EndoAtlas, features = "ProgesteroneSerum", raster = FALSE)
SFig1d
ggsave(filename = paste0(dir_out, 'SFig1d.pdf'),
       plot = SFig1d,
       width =20, height=9)

Supplementary Data Figure 1e

```{r}
dir_out <- paste0(dir_ENDO,"_Data/03_PCA_analysis/")

@@

Supplementary Data Figure 1f
CODE OK
"/home/common/data/output/projects/ENDO/E044/A082/SCTassay/"
```{r}

@@

Supplementary Data Figure 2
CODE OK
```{refined cluster markers}

@@

Supplementary Figure 3b and e
CODE running
```{r}
epithelial_cells_monocle3 <- readRDS(paste0(dir_data, "epithelial_cells_monocle3.rds"))

@@

Supplementary Data Figure 3c
Look in IBU folder for object with trajectory line or A063/A066
```{r}
#Figure S3c epithelial cells
FeaturePlot
Fig3c <- DimPlot(EndoAtlas_epithelial_reintegrated, reduction = "umap", group.by = "MenstrualCycleDay", raster = FALSE)
Fig3c
ggsave(filename = paste0(dir_out, 'Fig3c.pdf'),
       plot = Fig3c,
       width =20, height=9)
```
